---
title: U.S. Severe Weather Conditions with Highest Impact on Humans, Properties &
  Crops
author: "Andreas Manuth"
output:
  html_document:
    df_print: paged
---

# Synopsis
The highest impact on humans have weather event types with high death toll, as these come with high injuries, too. The top of this list are tornadoes, excessive heat, flash-/floods, heat, lightning, rip currents and thunderstorm winds. The highest impact on property and crop have tornadoes, flash-/floods and thunderstorm winds, which is a subset of those impacting humans directly.

# Introduction
The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events, esp.:
1. Across the United States, which types of events  are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?
The intended audience are government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, this report does not make any specific recommendations.

# Data Processing
As a standard tool set for data science the library tidyverse is loaded and setup to run before any chunk of code herein:
```{r setup}
library(tidyverse)
library(stringdist)
```

The analysis is based on the data from U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database made available as a comma-separated-value file compressed via the bzip2 algorithm to reduce its size at https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
Loading the first 1000 entries - unfortunately the current hardware can't deal with it all at once - for a first overview is done by:
```{r}
# Download 1st, because read_csv() does not un-compress remote bz2 files.
# The if-clauses are to prevent time consuming execution, when it is not needed.
if (!file.exists("repdata_data_StormData.csv.bz2"))
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
                  "repdata_data_StormData.csv.bz2") 
if (!exists("storm_data")) storm_data <- read_csv("repdata_data_StormData.csv.bz2", n_max = 1000)
summary(storm_data)
```
Some of the variables are defined at:  
* National Weather Service Storm Data Documentation https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
This describes the types of events and how the events are described in the data. Unfortunately a formal definition of the event properties is not given.  
* National Climatic Data Center Storm Events FAQ https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf provides  information about fatality codes, the Saffir-Simpson hurrican scale and some more details.
For completeness of the report repository they are downloaded as well:
```{r}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf", 
              "repdata_peer2_doc_pd01016005curr.pdf")
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf", 
              "repdata_peer2_doc_NCDC Storm Events-FAQ Page.pdf")
```
Now read all the data in the columns of interest and get the complete overview:
```{r}
storm_data <- read_csv("repdata_data_StormData.csv.bz2", 
                       col_types = cols_only(EVTYPE = col_character(), 
                                             FATALITIES = col_double(),
                                             INJURIES = col_double(),
                                             PROPDMG = col_double(),
                                             PROPDMGEXP = col_character(),
                                             CROPDMG = col_double(),
                                             CROPDMGEXP = col_character(), 
                                             # spec() says col_logical, but there are "k" etc.
                                             REFNUM = col_double()))
summary(storm_data)
```

According to descriptions the events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete. The higher weight of recent years given by that may also bring a focus on recent events taking account trends like climate heating.    
The event types in the data are, e.g.:
```{r}
event_types <- unique(storm_data$EVTYPE)
head(event_types)
tail(event_types)
print("Fatalities:")
summary(storm_data$FATALITIES)
print("Injuries:")
summary(storm_data$INJURIES)
```

The impact on population health is measured in terms of FATALITIES and INJURIES as numbers. The economic impact seem to be measured by PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP. As they are poorly documented we first look into the ...EXP variables:
```{r}
print(c("PROPDMGEXP values: ", unique(storm_data$PROPDMGEXP)))
print(c("CROPDMGEXP values: ", unique(storm_data$CROPDMGEXP)))
```
I assume that C/PROPDMGEXP is the exponent to the base of 10 of the values given in C/PROPDMG. Non-numeric characters are assumed to be interpreted like this:
- H / h = 2
- K = 3
- M / m = 6
- B = 9
- NA / + / ? / - = 0  
For the computation of property damages this interpretation is coded into a function and then added to the storm data:
```{r}
damage <- function(significand, order_of_magnitude){
    order_of_magnitude <- if (is.numeric(order_of_magnitude)) order_of_magnitude # no need to convert
    else # strange values are ignored
        if (is.na(order_of_magnitude) || str_detect(as.character(order_of_magnitude), "[:punct:]"))
            order_of_magnitude <- 0
        else { # handling of common abbreviations
            order_of_magnitude <- str_to_lower(order_of_magnitude)
            if (order_of_magnitude == "h") order_of_magnitude <- 2
            else if (order_of_magnitude == "k") order_of_magnitude <- 3
            else if (order_of_magnitude == "m") order_of_magnitude <- 6
            else if (order_of_magnitude == "b") order_of_magnitude <- 9
            else order_of_magnitude <- NaN # unrecognized characters result in strange values
        }
    significand * 10^order_of_magnitude
}
storm_data <- mutate(storm_data, 
                     propdmg_unified = damage(PROPDMG, PROPDMGEXP),
                     cropdmg_unified = damage(CROPDMG, CROPDMGEXP))
print("Property Damage:")
summary(storm_data$propdmg_unified)
print("Crop Damage:")
summary(storm_data$cropdmg_unified)
```
So maximum property damage is `r max(storm_data$propdmg_unified)` and for crops it is `r max(storm_data$propdmg_unified)`. Now make sure event types don't differ by upper/lower case and group the data by event type and picture the impact by event type:
```{r}
storm_data <- mutate(storm_data, EVTYPE = str_to_upper(EVTYPE))
event_damage <- group_by(storm_data, EVTYPE)
totals <- summarize(event_damage, 
                    fatalities = sum(FATALITIES), 
                    injuries = sum(INJURIES), 
                    propdmg = sum(propdmg_unified), 
                    cropdmg = sum(cropdmg_unified))
summary(totals)
head(totals)
tail(totals)
```
The tail end shows that there are event types without any damage, which are extracted for concise summary and further analysis of the harmful types: 
```{r}
harmless_events <- select(filter(totals, 
                                 fatalities == 0 & injuries == 0 & propdmg == 0 & cropdmg == 0), 
                          EVTYPE)
harmful_events <- filter(totals, fatalities != 0 | injuries != 0 | propdmg != 0 | cropdmg != 0)
summary(harmful_events)
head(harmful_events, 10)
```
The first five rows (default head()) showed typo "AVALNCE" and therefore separated from the following (displaying 10 rows) "AVALANCHE". So the EVTYPEs needs to be standardized in terms of spelling and then streamlined again. For standardization use the "soundex" algorithm encoded into the phonetic() function of the stringdist library:
```{r}
harmful_events <- mutate(harmful_events, event_phonetic = phonetic(EVTYPE))
harmful_phonetic<- group_by(harmful_events, event_phonetic)
harmful_totals <- summarize(harmful_phonetic, 
                            event_type = EVTYPE[1], # readable, but spelling may be wrong
                            fatalities = sum(fatalities), 
                            injuries = sum(injuries), 
                            propdmg = sum(propdmg), 
                    cropdmg = sum(cropdmg))
summary(harmful_totals)
head(harmful_totals)
tail(harmful_totals)
```
# Results

## Human Harm
There are `r nrow(filter(harmful_totals, fatalities != 0))` event types that caused deaths and `r nrow(filter(harmful_totals, injuries != 0))` that caused injuries. These are plotted here on logarithmic scales (TORNADO is much bigger than the others) and then listed sorted descending first by deaths than by injured:
```{r}
humans_harmed <- arrange(filter(harmful_totals, fatalities != 0 | injuries != 0), 
                         desc(fatalities), desc(injuries))
ggplot(humans_harmed, aes(fatalities, injuries)) + 
    geom_text(aes(label = event_type)) + scale_x_log10() + scale_y_log10() + labs(title = "Humans Harmed")
print(humans_harmed)
```
https://www.severe-weather.eu/severe-weather-outlooks-faq/ says TSTM stands for thunderstorm.
As the event types cluster around the diagonal it seems sensible to concentrate on event types with high death toll as these come with high injuries, too.

## Damage
There are `r nrow(filter(harmful_totals, propdmg != 0))` that caused property damage and `r nrow(filter(harmful_totals, cropdmg != 0))` that caused crop damage. These are plotted below and listed sorted descending first by damage to property than by crop:
```{r}
cash_dmg <- arrange(filter(harmful_totals, propdmg != 0 | cropdmg != 0), 
                         desc(propdmg), desc(cropdmg))
ggplot(cash_dmg, aes(propdmg, cropdmg)) + 
    geom_text(aes(label = event_type))  + labs(title = "Property & Crop Damage")
print(cash_dmg)
```
Clearly the first six on the list have the biggest impact on property and crops. They are a subset of the top types harmful to humans.