---
title: "U.S. storm population harm and economic impact"
author: "Andreas Manuth"
output: html_notebook
---
#Synopsis


#Introduction
The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events, esp.:
1. Across the United States, which types of events  are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?
The intended audience are government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, this report does not make any specific recommendations.

#Data Processing
The analysis is based on the data from U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database made available as a comma-separated-value file compressed via the bzip2 algorithm to reduce its size at https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
Loading it for processing and a first overview is done by
```{r}
library(tidyverse)
# download 1st, because read_csv() does not un-compress remote bz2 files
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
              "repdata_data_StormData.csv.bz2") 
storm_data <- read_csv("repdata_data_StormData.csv.bz2")
summary(storm_data)
```
Some of the variables are defined at:
- National Weather Service Storm Data Documentation https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
- National Climatic Data Center Storm Events FAQ https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf
For completeness of the report repository they are downloaded as well:
```{r}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf", 
              "repdata_peer2_doc_pd01016005curr.pdf")
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf", 
              "repdata_peer2_doc_NCDC_Storm_Events-FAQ_Page.pdf")
```

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

#Results